---
title: "Case Study 1"
subtitle: "AKSTA Statistical Computing"
output: pdf_document
header-includes:
  - \usepackage{booktabs}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*The .Rmd and .html (alternatively the .pdf) should be uploaded in TUWEL
by the deadline. Refrain from using explanatory comments in the R code
chunks but write them as text instead. Please consider the AI Guidelines
as specified in the organizational slides (you can use the table).
Points will be deducted if the submitted file is not in a decent form;
answers should be answering the concrete questions and not discuss
topics in general - please be mindful of our tutors' time and the size
of the course.*

\begin{table}[ht]
\centering
\caption{You can use and extend this table to specify your AI usage (if used at all).}
\begin{tabular}{l|rrr}
\toprule
 \textbf{AI usage} & \textbf{Task 1} & \textbf{Task 2} &  \ldots \\
\midrule
\textit{Correcting faulty code}   &  no  &  &    \\
\textit{Improving existing code}   &  little   &  &     \\
\textit{Generating new code}   &  medium   &  &    \\
\textit{Layout and formatting}  &  heavy   &  &     \\
\textit{Text correction}   &    &  &    \\
 \ldots  &    &  &    \\
\bottomrule
\end{tabular}
\end{table}

# 1. Ratio of Fibonacci numbers (1 point)

## a.

Write two different R functions which return the sequence
$r_{i}=F_{i+1}/F_i$ for $i=1,\ldots,n$ where $F_i$ is the $i$th
Fibonacci number, once using `for` and once using `while`.

```{r}
fibonacci_for_loop <- function(n){
  F <- numeric(n + 1)
  F[1] <- 1
  F[2] <- 1
  
  if (n >= 2) {
    for (i in 3:(n + 1)) {
      F[i] <- F[i - 1] + F[i - 2]
    }
  }
  
  r <- numeric(n)
  for (i in 1:n) {
    r[i] <- F[i + 1] / F[i]
  }
  
  r
}

fibonacci_while_loop <- function(n){
  F <- numeric(n + 1)
  F[1] <- 1
  F[2] <- 1
  
  i <- 3
  while (i <= n + 1) {
    F[i] <- F[i - 1] + F[i - 2]
    i <- i + 1
  }
  
  r <- numeric(n)
  i <- 1
  while (i <= n) {
    r[i] <- F[i + 1] / F[i]
    i <- i + 1
  }
  r
}
```

## b.

Benchmark the two functions for $n=200$ and $n=2000$ (you can use
package **microbenchmark** or package **bench** for this purpose). Which
function is faster?

```{r}
library(microbenchmark)

microbenchmark(
  fibonacci_for_loop(200),
  fibonacci_while_loop(200),
  times = 100
)

microbenchmark(
  fibonacci_for_loop(2000),
  fibonacci_while_loop(2000),
  times = 100
)
```

**Conclusion:**

\par

The for loop is for the ratio of Fibonacci numbers overall at n=200 and
n=2000 faster than the while loop.

## c.

Plot the sequence for $n=100$. For which $n$ does it start to stabilize
to a value? What number does the sequence converge to?

```{r}
r100 <- fibonacci_for_loop(100)

plot(
  1:100, 
  r100, 
  type = "b", 
  pch = 16,
  xlab = "n", 
  ylab = expression(r[n]),
  main = "Ratio of Fibonacci numbers"
)

# golden ratio
abline(h = (1 + sqrt(5)) / 2, lty = 2)
```

**Conclusion:**

\par

For small n the sequence fluctuates, but stabilizes quite fast at around
n=\~10 and it it is known, that the Fibonacci sequence converges to the
golden ratio:

$$
\varphi = \frac{1+\sqrt{5}}{2}
$$

# 2. Gamma function (1 point)

## a.

Write a function to compute the following for $n$ a positive integer
using the `gamma` function in base R. $$
\rho_n =\frac{\Gamma((n-1)/2)}{\Gamma(1/2)\Gamma((n - 2)/2)}
$$

```{r}
gamma_compute <- function(n){
  rho <- gamma((n-1)/2) / (gamma(1/2) * gamma((n-2)/2))
  return(rho)
}
```

## b.

Try $n = 2000$. What do you observe? Why do you think the observed
behavior happens?

```{r}
gamma_compute(2000)
```

We see that the call to the function returns "Not a Number". This is
because `gamma(1000)=1000!` is significantly larger than the largest
number R can handle, therefore the function tires to devide infinity by
infinity, which returns "NaN"!

## c.

Write an implementation which can also deal with large values of
$n>1000$.

In order to do that, we use `lgamma()`, which returns the logarithm of
the gamma function, which keeps the numbers involved smaller and turns
the division in to a subtraction. At the end we exponentiate everything
in order to get the true value:

```{r}
gamma_compute_large <- function(n){
  log_result <-lgamma((n - 1) / 2) - lgamma(1 / 2) - lgamma((n - 2) / 2)
  return(exp(log_result))
}

gamma_compute_large(2000)
```

## d.

Plot $\rho_n/\sqrt{n}$ for different values of $n$. Can you guess the
limit of $\rho_n/\sqrt{n}$ for $n \rightarrow\infty$?

```{r}
N <- seq(1,2000, by = 10)

plot(N, gamma_compute_large(N)/sqrt(N), type = "l", col = "red", ylab = expression(rho[n]/sqrt(n)), xlab = "n", main = "Asymptotic limit")
#curve(log(x), add = TRUE, col="blue")
```

It seems to converge to approximately $0.4$.

# 3. Weak law of large numbers (1.5 points)

The law of large numbers states that if you repeat an experiment
independently a large number of times and average the result, what you
obtain should be close to the expected value. Formally, the weak law of
large numbers states that if $X_1, X_2, \ldots X_n$ are independently
and identically distributed (iid) random variables with a finite
expected value $E(X_i)=\mu<\infty$, then, for any $\epsilon > 0$
$\lim_{n\rightarrow \infty} P(|\bar X_n-\mu|>\epsilon)=0$ where
$\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$ is the sample mean (the sample
mean converges in probability to the expected value).

## a.

-   In R, simulate rolling a fair 6-sided die $n=10\,000$ times using
    vectorized functions.

-   After each roll, compute the cumulative mean of all previous rolls.

-   Plot the running average against the number of rolls and comment on
    the result.

## b.

-   Replicate the die-rolling experiment $M=50$ times using a for loop.
    This will result in $50$ paths which you can plot against the number
    of rolls.

-   Plot all 50 paths on the same graph to observe the variation in the
    running averages.

## c.

What is the theoretical expected value of a fair 6-sided die?

## d.

Comment on how the running mean of the paths behaves relative to the
expected value as $n$ increases.

## e.

-   Modify the code in b. to use a while loop.

-   Stop the loop when the proportion of replications where the
    difference between the running average and the expected value lies
    outside the interval $[-\epsilon,\epsilon]$ is less than 0.05. Use
    $\epsilon = 0.01$.

-   How high is $n$ at the end of the loop?

## f.

gitgit

-   Repeat the analysis in b, but instead of simulating from a discrete
    uniform distribution, simulate from a continuous standard cauchy
    distribution using `rcauchy()`. The (standard) cauchy distribution
    arises from the ratio of two independent mean zero (standard)
    normally distributed random variables.

-   Plot the running mean against increasing $n$. Comment on whether the
    weak Law of Large Numbers seems to hold for this distribution.
    Explain why or why not.

# 4. Central limit theorem (1.5 points)

The central limit theorem states that, under certain conditions, the sum
of a large number of random variables is approximately normal. Consider
the case where $X_1, X_2,\ldots,X_n$ are i.i.d. random variables with
finite expected values ($E(X_i)=\mu<\infty$) and variances
$(Var(X_i)=\sigma^2<\infty)$. The central limit theorem states that the
cumulative distribution function (CDF) of the normalized random variable
$$
Z_n = \frac{\bar X_n - \mu}{\sigma/\sqrt{n}}
$$ converges to the standard normal CDF as $n\rightarrow\infty$.

## a.

Derive the expected value of the sample mean
$\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$.

**Conclusion:**

$\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n} = \frac{1}{n}\sum_{i=1}^nX_i$

1.  Apply expected value operator:

    $\mathbb{E}(\bar{X}_n) = \mathbb{E}(\frac{1}{n}\sum_{i=1}^n X_i)$

2.  Use linearity property of E operator:

    $\mathbb{E}(\bar{X}_n) = \frac{1}{n}(\sum_{i=1}^n \mathbb{E}(X_i))$

3.  Use i.i.d properties

    Due to $X_i$ being i.i.d $\Rightarrow$

    $\mathbb{E}(X_i)=\mu$

    $\mathbb{E}(X_n) = \frac{1}{n}\sum_{i=1}^n \mu$

4.  Calculate sum

    $\mathbb{E}(X_i) = \frac{1}{n}\cdot n\mu = \mu$

5.  Result:

    $\mathbb{E}(\bar X_n) = \mu$

## b.

Derive the variance of the sample mean
$\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n}$.

**Conclusion:**

$\bar X_n=\frac{X_1+X_2+\ldots+X_n}{n} = \frac{1}{n}\sum_{i=1}^nX_i$

1.  Apply Variance operator:

    $\mathrm{Var}(\bar{X}_n) = \mathrm{Var}\left(\frac{1}{n}\sum_{i=1}^n X_i\right)$

2.  Factorize:

    $\mathrm{Var}(\bar X_n) = \frac{1}{n^2} \mathrm{Var}\left(\sum_{i=1}^n X_i\right)$

3.  Apply independence:

    $\mathrm{Var}(\bar X_n) = \frac{1}{n^2} \sum_{i=1}^n \mathrm{Var}(X_i)$

4.  Due to $X_i$ being i.i.d $\Rightarrow$

    $\mathrm{Var}(\bar X_i) = \sigma^2$

    $\mathrm{Var}(\bar X_n) = \frac{1}{n^2} \sum_{i=1}^n \sigma^2$

5.  Calculate sum

    $\mathrm{Var}(\bar X_i) = \frac{1}{n^2} \cdot n \sigma^2$

6.  Result:

    $\mathrm{Var}(\bar X_n) = \frac{\sigma^2}{n}$

## c.

The central limit theorem makes few assumption on the distribution of
the random variables $X_i$. Consider again the die-rolling experiment
where $X_i$ is the random variable giving the number in roll $i$.

-   Simulate Die Rolls: For each value of $n$ (where
    $n=\{100,1000,10000\}$), simulate rolling a die $n$ times. Record
    the mean of the $n$ rolls.

-   Repeat the above process $M=100$ times for each value of $n$. This
    will give you 100 sample means for each $n$.

-   Plot a histogram of the 100 sample means for each value of $n$.
    Observe the shape of the histograms.

-   Discuss how the distribution of the sample means changes as $n$
    increases. Comment on whether the histograms appear to approach a
    normal distribution as $n$, illustrating the Central Limit Theorem.

```{r, fig.width=10}
set.seed(12318680)

n.values <- c(100, 1000, 10000)
M <- 100

sample_means <- vector("list", length(n.values))
names(sample_means) <- paste0("n=", n.values)
par(mfrow = c(1, 3))
for (j in seq_along(n.values)) {
  n <- n.values[j]
  
  rolls_mean <- numeric(M)
  
  for (m in 1:M) {
    rolls <- sample(1:6, size = n, replace = TRUE)
    rolls_mean[m] <- mean(rolls)
  }
  hist(
    rolls_mean,
    main = paste("Histogram of sample means\nn =", n.values[j]),
    xlab = expression(bar(X)[n]),
    col = "lightgray",
    border = "white"
  )
}
```
**Conclusion:** \par
- **n=100:**the distribution is relatively spread out and a bell-shape is visible \par
- **n=1000:**we can see with a bigger sample means size, the distribution is more concentrated around the mean and appears more symmetric
- **n=10000:** we can see, that the sample means are tightly clustered around $\mu = 3.5$ and the histogram seems to follow a normal distribution

To conclude, as n increases the variance of the sample mean decreases and the values are centered around the expected value and the distribution has the shape of a normal distribution. This behaviour is according to the central limit theorem.

## d.

Compute expected value $\mu$ and the variance $\sigma^2$ for the random
variable $X_i$ (giving the number in die-roll $i$), which follows a
discrete uniform distribution.

## e.

Now that you computed $\mu$ and $\sigma^2$, consider $Z_n$.

-   For each value of $n$ (where $n=\{100,1000,10000\}$) and for
    $M=\{100,1000,10000\}$ replications, simulate the rolling of $n$
    dice and calculate the $Z_n$.

-   Plot histograms of $Z_n$ for each combination of $n$ and $M$.

-   Discuss how the distribution of $Z_n$ changes as $n$ and $M$
    increase. Comment on whether the distributions of $Z_n$ appear to
    approach a normal distribution, illustrating the Central Limit
    Theorem.

# 5. Readable and efficient code (1 points)

Read over the code below.

## a.

Explain (in text) what the code does.

## b.

Explain (in text) what you would change to make the code more readable.

## c.

Change the code according to b. and wrap it in a function. This function
should have at most 10 lines (without adding commands to more lines such
as `x <- 1; y <- 2`. Such commands will count as 2 lines!). Check that
the function called on the same input outputs the same as the provided
code.

```{r eval=FALSE}
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)
if (sum(x >= .001) < 1) {
  stop("step 1 requires 1 observation(s) with value >= .001")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- sin(r) + .01
if (sum(x >= .002) < 2) {
  stop("step 2 requires 2 observation(s) with value >= .002")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 2 * sin(r) + .02
if (sum(x >= .003) < 3) {
  stop("step 3 requires 3 observation(s) with value >= .003")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 3 * sin(r) + .03
if (sum(x >= .004) < 4) {
  stop("step 4 requires 4 observation(s) with value >= .004")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 4 * sin(r) + .04
x
```

```{r eval=FALSE}
set.seed(1) 
x <- rnorm(1000)
y <- 2 + x + rnorm(1000)
df <- data.frame(x, y)

cat("Step", 1, "\n")
fit1 <- lm(y ~ x, data = df[-(1:250),])
p1 <- predict(fit1, newdata = df[(1:250),])
r <- sqrt(mean((p1 - df[(1:250),"y"])^2))  

cat("Step", 2, "\n")
fit2 <- lm(y ~ x, data = df[-(251:500),])
p2 <- predict(fit2, newdata = df[(251:500),])
r <- c(r, sqrt(mean((p2 - df[(251:500),"y"])^2)))

cat("Step", 3, "\n")
fit3 <- lm(y ~ x, data = df[-(501:750),])
p3 <- predict(fit3, newdata = df[(501:750),])
r <- c(r, sqrt(mean((p3 - df[(501:750),"y"])^2)))

cat("Step", 4, "\n")
fit4 <- lm(y ~ x, data = df[-(751:1000),])
p4 <- predict(fit4, newdata = df[(751:1000),])
r <- c(r, sqrt(mean((p4 - df[(751:1000),"y"])^2)))  
r
```

# 6. Measuring and improving performance (2 points)

Have a look at the code of the function below. It is a function for
performing a Kruskal-Wallis test, a robust non-parametric method for
testing whether samples come from the same distribution. (Note: we
assume no missing values are present in `x`).

```{r}
kwtest <- function (x, g, ...) 
{
    if (is.list(x)) {
        if (length(x) < 2L) 
            stop("'x' must be a list with at least 2 elements")
        if (!missing(g)) 
            warning("'x' is a list, so ignoring argument 'g'")
        if (!all(sapply(x, is.numeric))) 
            warning("some elements of 'x' are not numeric and will be coerced to numeric")
        k <- length(x)
        l <- lengths(x)
        if (any(l == 0L)) 
            stop("all groups must contain data")
        g <- factor(rep.int(seq_len(k), l))
        x <- unlist(x)
    }
    else {
        if (length(x) != length(g)) 
            stop("'x' and 'g' must have the same length")
        g <- factor(g)
        k <- nlevels(g)
        if (k < 2L) 
            stop("all observations are in the same group")
    }
    n <- length(x)
    if (n < 2L) 
        stop("not enough observations")
    r <- rank(x)
    TIES <- table(x)
    STATISTIC <- sum(tapply(r, g, sum)^2/tapply(r, g, length))
    STATISTIC <- ((12 * STATISTIC/(n * (n + 1)) - 3 * (n + 1))/(1 - 
        sum(TIES^3 - TIES)/(n^3 - n)))
    PARAMETER <- k - 1L
    PVAL <- pchisq(STATISTIC, PARAMETER, lower.tail = FALSE)
    names(STATISTIC) <- "Kruskal-Wallis chi-squared"
    names(PARAMETER) <- "df"
    RVAL <- list(statistic = STATISTIC, parameter = PARAMETER, 
        p.value = PVAL, method = "Kruskal-Wallis rank sum test")
    return(RVAL)
}
```

## a.

Write a pseudo code outlining what the function does.

*Input:* dataset $x$ (either list or vector) and grouping factor $g$
(optional when x is a list)

1.  *Input Validation and Transformation* check if $x$ is a list

-   if yest: check if it has at least 2 elements.If not, stop. If yes
    proceed to create the grouping factor $g$, each element of the list
    representing one group. Then, transform the list to a single numeric
    vector.
-   if no: check if the total number of observations $n = |x|$ equals
    the length of $g$. If not, stop. If yes, transform $g$ to a factor
    and check if in each level there at least 2 observations. If not,
    stop

2.  *Rank Computation and Tie Identification*

-   Compute the ranks $r = (r_1, r_2, \dots, r_n)$ for each observation
    in $x$, assigning the average rank to tied values.
-   Construct a frequency table $T$ containing the counts of each unique
    value in $x$ to identify tied observations.

3.  *Test Statistic Calculation*

-   For each group $i \in \{1, \dots, k\}$, compute the sum of ranks
    $R_i = \sum_{j \in \text{group } i} r_j$ and let $n_i$ be the number
    of observations in group $i$.

-   Compute the unadjusted Kruskal-Wallis statistic $S$:

    $$S = \sum_{i=1}^k \frac{R_i^2}{n_i}$$

-   Apply the tie-correction factor to compute the final Kruskal-Wallis
    test statistic $H$:

    $$H=\frac{\frac{12}{n(n+1)}S-3(n+1)}{1-\frac{\sum_{t \in T}(t^3-t)}{n^3-n}}$$

4.  *P-value Computation*

-   Calculate the degrees of freedom $v = k - 1$.

-   Calculate the $p$-value using the right tail of the Chi-squared CDF
    with $v$ degrees of freedom:

    $$p\text{-value} = 1 - F_{\chi^2_v}(H)$$

*Output:* A list containing the test statistic H, the degrees of freedom
$v$ and the p-value.

## b.

For example data, call the function in two ways: once using `x` as a
list and once using `x` as a vector with a corresponding `g` argument.
Ensure that the two different function calls return the same thing by
aligning the inputs.

```{r}
set.seed(1899)
sample1 = rnorm(100, 4,1)
sample2 = rnorm(100,5,1)
sample3 = rnorm(100,5,2)
x_list <- list(sample1 = sample1, sample2 = sample2, sample3 = sample3)
x_vec <- c(sample1, sample2, sample3)
g <- rep(1:3, each = 100)
kwtest(x_list)
kwtest(x_vec, g)
```

We see, the results are identical.

## c.

Make a faster version of `kwtest()` that only computes the
Kruskal-Wallis **test statistic** when the input is a numeric variable
`x` and a variable `g` which gives the group membership. You can try
simplifying the function above or by coding from the mathematical
definition (see
<https://en.wikipedia.org/wiki/Kruskal%E2%80%93Wallis_one-way_analysis_of_variance>).
This function should also perform some checks to ensure the correctness
of the inputs (use `kwtest()` as inspiration).

```{r}
kw_fast <- function(x, g) {
  n <- length(x)
  if (!is.numeric(x)) stop("x must be a numeric vector!")
  if (n != length(g)) stop("'x' and 'g' must have the same length")
  if (n < 2L) stop("not enough observations")
  g <- as.factor(g)
  k <- nlevels(g)
  if (k < 2L) stop("all observations are in the same group")
  r <- rank(x)
  rank_sums <- rowsum(r, g)
  group_lengths <- tabulate(g)
  STAT <- sum(rank_sums^2 / group_lengths)
  TIES <- table(x)
  STAT <- ((12 * STAT / (n * (n + 1)) - 3 * (n + 1)) / 
          (1 - sum(TIES^3 - TIES) / (n^3 - n)))
  names(STAT) <- "Kruskal-Wallis chi-squared"
  return(STAT)
}
kw_fast(x_vec, g)
```

We see our new function returns the same test statistic as in part b. We
use optimized functions for counting: `rowsum()` and `tabulate()`
instead of the slower but more flexible `tapply()`.

## d.

Consider the following scenario. You have samples available from
multiple experiments $m=1000$ where you collect the numerical values for
the quantity of interest `x` and the group membership for $n$
individuals. The first $20$ individuals in each sample belong to group
1, the following $20$ individuals in each sample belong to group 2, the
last $10$ individuals in each sample belong to group 3. Use the
following code to simulate such a data structure:

```{r}
set.seed(1234)
m <- 1000 # number of repetitions
n <- 50   # number of individuals
X <- matrix(rt(m * n, df = 10), nrow = m)
grp <- rep(1:3, c(20, 20, 10))
```

Write a function which performs the Kruskal-Wallis test using the
function `stats:::kruskal.test.default()` for $m$ repeated experiments
and returns a vector of $m$ test statistics. The input of this function
are a matrix `X` with $m$ rows and $n$ columns and a vector `g` which
gives the grouping.

```{r}
kw_default <- function(X,g) {
  apply(X, 1, function(y) {stats:::kruskal.test.default(y, g)$statistic})
}
head(kw_default(X,grp))
```

## e.

Write a function which performs the Kruskal-Wallis test using the
function in point c. for $m$ repeated experiments and returns a vector
of $m$ test statistics. The input of this function are a matrix `X` with
$m$ rows and $n$ columns and a vector `g` which gives the grouping.

```{r}
kw_fast_multiple <- function(X,g) {
  apply(X, 1, function(y) {kw_fast(y,g)})
}
head(kw_fast_multiple(X,grp))
```

The to implementations return the same test statistics, as expected.

## f.

Compare the performance of the two approaches using a benchmarking
package on the data generated above. Comment on the results.

```{r}
library(microbenchmark)

bench_results <- microbenchmark(default_R = kw_default(X,grp),
                                fast_custom = kw_fast_multiple(X, grp),
                                times = 10)
print(bench_results)
```

Our custom implementation is indeed significantly faster than the base R
implementation (about 70 milliseconds on average). This is because we
used highly optimized functions and skip the formatting and p-value
computation.

## g.

Now consider vectorizing the function in point c. Compare this approach
to the other two. Comment on the results.

```{r}
kw_fast_vector <- function(X,g) {
  if (!is.matrix(X) || !is.numeric(X)) stop("X must be a numeric matrix!")
  n <- ncol(X)
  if (n != length(g)) stop("The number of columns in 'X' must equal the length of 'g'")
  if (n < 2L) stop("not enough observations")
  
  g <- as.factor(g)
  if (nlevels(g) < 2L) stop("all observations are in the same group")
  
  R <- t(apply(X, 1, rank))
  
  D <- model.matrix(~ 0 + g) 
  rank_sums <- R %*% D
  
  group_lengths <- tabulate(g)
  STAT_base <- rowSums(sweep(rank_sums^2, 2, group_lengths, "/"))
  H_unadj <- (12 * STAT_base / (n * (n + 1))) - 3 * (n + 1)
  
  tie_sums <- apply(X, 1, function(row_data) {
    if (anyDuplicated(row_data)) {
      ties <- tabulate(match(row_data, unique(row_data)))
      sum(ties^3 - ties)
    } else {
      0 
    }
  })

  tie_correction <- 1 - tie_sums / (n^3 - n)
  
  STAT <- H_unadj / tie_correction
  
  return(STAT)
}
head(kw_fast_vector(X,grp))

bench_results_2 <- microbenchmark(default_R = kw_default(X,grp),
                                fast_custom = kw_fast_multiple(X, grp),
                                fast_custom_vec = kw_fast_vector(X, grp),
                                times = 10)
print(bench_results_2)
```

This implementation is by far the fastest, since R handles vectorized
operation very well this is not a surprise (the extreme difference
however was a surprise).
