---
title: "Case Study 3"
author: "Oliver Crasnianec, Lorenz Kovsca, Nikolai Asche"
date: "2026-05-06"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

*The .Rmd* **and** *.html (or .pdf) should be uploaded in TUWEL by the
deadline. Refrain from using explanatory comments in the R code chunks
but write them as text instead. Points will be deducted if the submitted
file is not in a decent form.*

# Data

Load the data set you exported in the final Task of Case Study 2.
Eliminate all observations with missing values in the income status
variable.

As a reminder, the data set includes world data from 2020, focusing on:

- **Education Expenditure (% of GDP)**
- **Youth Unemployment Rate (15-24 years)**
- **Net Migration Rate** (difference between the number of people
  entering and leaving a country per 1,000 persons)

for most world entities in 2020. The data was downloaded from
<https://www.cia.gov/the-world-factbook/about/archives/>. Additional
information on continent, subcontinent/region and income status was
appended to the dataset in Case Study 2.

```{r}
library(tidyverse)
library(forcats)


data_raw <- read_csv("file_out2.csv",
                     na = c(".", "NA", ""),
                     show_col_types = FALSE)

data <- data_raw %>%
  rename(income_status = status) %>%
  filter(!is.na(income_status)) %>%
  mutate(
    income_status = factor(income_status, levels = c("H", "UM", "LM", "L")),
    continent = factor(continent),
    subcontinent = factor(subcontinent)
  )

glimpse(data)

table(data$income_status)
```

**Conclusion:**

We used the provided file "file_out2.csv" provided in the TUWEL section
of Assignment 3 and renamed status to income_status

# Tasks:

## a. Education expenditure in different income levels

Using **ggplot2**, create a density plot of the education expenditure
grouped by income status. The densities for the different groups are
superimposed in the same plot rather than in different plots. Ensure
that you order the levels of the income status such that in the plots
the legend is ordered from High (H) to Low (L).

- The color of the density lines is black.

- The area under the density curve should be colored differently among
  the income status levels.

- For the colors, choose a transparency level of 0.5 for better
  visibility.

- Position the legend at the top center of the plot and give it no title
  (hint: use `element_blank()`).

- Rename the x axis as "Education expenditure (% of GDP)"

Comment briefly on the plot.

```{r}
ggplot(data, aes(x = expenditure, fill = income_status)) +
  geom_density(color = "black", alpha = 0.5, na.rm = TRUE) +
  labs(
    x = "Education expenditure (% of GDP)",
    y = "Density",
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

**Conclusion:**

By inspecting the plot, we can see, that the education expenditure (% of
GDP) strongly overlaps across all 4 income status groups e.g. across H,
UM, LM and L. Most countries seem to spend roughly between 3% and 6% of
GDP on education, so there seems not to be a clear separation between
the income levels. Higher income countries seem to concentrate more
around 5% of GDP, meanwhile lower income countries seem to gravitate
more towards lower education expenditure (% of GDP). Overall, the
differences between the income groups are not very pronounced.

## b. Income status in different continents

Investigate how the income status is distributed in the different
continents.

- Using **ggplot2**, create a stacked barplot of absolute frequencies
  showing how the entities are split into continents and income status.
  Comment the plot.

```{r}
ggplot(data, aes(x = continent, fill = income_status)) + geom_bar() +
  labs(x = "Continents", 
       y = "Absolute frequency", 
       fill = "Income status") +
  theme_minimal() +
  theme(legend.position = "top")
```

**Conclusion:**

The absolute frequency stacked barplots show that the number of entities
differs strongly across continents. Africa and Asia contain many lower-
and lower-middle income entities, while Europe seems to have the most
high income entities. Asia and the Americas have a more mixed income
distribution. Further Oceania seems to have fewer observations overall,
as it's stacked barplot is compared to the others rather small.

- Create another stacked barplot of relative frequencies (height of the
  bars should be one). Comment the plot.

```{r}
ggplot(data, aes(x = continent, fill = income_status)) + geom_bar(position = "fill") +
  scale_y_continuous(labels = scales:: percent) +
  labs(x = "Continents", 
       y = "Relative frequency", 
       fill = "Income status") +
  theme_minimal() +
  theme(legend.position = "top")
```

**Conclusion:**

The stacked barplot of relative frequency makes the compositions of
incomes across continents easier to compare, as Oceania is now easier to
compare to the others. Europe has the largest proportion of high income
entities, while Africa has the largest proportion of low- and
lower-middle income entities. Asia and the Americas seem to be more
mixed. Oceania mainly consists of lower-middle, upper-middle and
high-income entities with no low-income entities.

- Create a mosaic plot of continents and income status using base R
  functions.

```{r}
income_continents <- table(data$continent, data$income_status)
income_status_colors <- c(
  "H" = "#f8766d",
  "UM" = "#7cae00",
  "LM" = "#00bfc4",
  "L" = "#c77cff"
)
par(
  mar = c(5, 4, 4, 8),
  xpd = TRUE
)

mosaicplot(
  income_continents,
  color = income_status_colors,
  main = "Income status by continent",
  xlab = "Continent",
  ylab = "Income status"
)

legend(
  "topright",
  inset = c(-0.25, 0),
  legend = names(income_status_colors),
  fill = income_status_colors,
  title = "Income status",
  bty = "n",
  cex = 0.8
)
```

**Conclusion:**

By using a mosaic plot, we obtain a useful visual representation and
overview of the income status of each continent and it is possible to
recognize relationships between different variables. The width shows the
number of observations per continent and the height of each colored part
shows the income distribution within each continent.

- Briefly comment on the differences between the three plots generated
  to investigate the income distribution among the different continents.

**Conclusion:**

The absolute frequency stacked barplot is useful for comparing the total
number in each continent, but in turn makes it harder to compare
proportions as in the case of Oceania having a lower number of
observations compared to the other continents. By using the relative
frequency stacked barplot it is possible to remove the effect the number
of observations of each continent have on size of each bar and therefore
making income composition more clearly visible. Lastly the mosaic plot
combines both features as mentioned before the width shows the number of
observations, while the height of each colored part shows the income
distribution within each continent.

## c. Income status in different subcontinents

For Oceania, investigate further how the income status distribution is
in the different subcontinents. Use one of the plots in b. for this
purpose. Comment on the results.

```{r}
oceania_data <- data %>%
  filter(continent == "Oceania") %>%
  mutate(subcontinent = fct_drop(subcontinent))

ggplot(oceania_data, aes(x = subcontinent, fill = income_status)) +
  geom_bar(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "Subcontinents Oceania",
    y = "Relative frequency",
    fill = "Income status"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

**Conclusion:**

I used the relative frequency stacked barblot from subtask b. for
further investigation, due to the subcontinents of Oceania having rather
different numbers of observation. By doing that it is easier to compare
proportions of income status between the subcontinents. We can see, that
Australia and New Zealand only consist of high-income entities.
Melanesia has the highest proportion of lower-middle-income entities,
meanwhile Polynesia has the highest proportion of upper-middle income
entities. Micronesia also seems to be more mixed but still with a high
proportion of high-income entities. Also there are no low-income
entities shown for Oceania in the data set.

## d. Net migration in different continents

- Using **ggplot2**, create parallel boxplots showing the distribution
  of the net migration rate in the different continents.

- Prettify the plot (change y-, x-axis labels, etc).

- Identify which country in Asia constitutes the largest negative
  outlier and which country in Asia constitutes the largest positive
  outlier.

- Comment on the plot.

```{r}
# find largest and smallest outliers in Asia
outliers_df <- data %>% filter(continent == "Asia") %>% 
                summarise(max_out = country[which.max(net_migr_rate)],
                          max_val = net_migr_rate[which.max(net_migr_rate)],
                          min_out = country[which.min(net_migr_rate)],
                          min_val = net_migr_rate[which.min(net_migr_rate)])

ggplot(data, aes(x = continent, y = net_migr_rate, fill = continent)) +
  geom_boxplot() +
  # annotate outliers in the Boxplot directly
  annotate("text", x = "Asia", y = outliers_df$max_val, 
           label = paste("          ",outliers_df$max_out)) +
  annotate("text", x = "Asia", y = outliers_df$min_val, 
           label = paste("                ",outliers_df$min_out)) +
  labs(title = "Boxplots of net migration rates",
       xlab = "Continent", ylab = "net migration rate")
```

**Conclusion:**

The largest positive outlier in Asia is Syria, the smallest Lebanon.
Apart from Oceania all continents have distributions with median roughly
at 0. Except for the extreme cases outliers roughly lie within +/- 15 of
the median.

## e. Net migration in different subcontinents

The graph in d. clearly does not convey the whole picture. It would be
interesting also to look at the subcontinents, as it is likely that a
lot of migration flows happen within the continent.

- Investigate the net migration in different subcontinents using again
  parallel boxplots. Group the boxplots by continent (hint: use
  `facet_grid` with `scales = "free_x"`).

- Remember to prettify the plot (rotate axis labels if needed).

- Describe what you see.

```{r}
ggplot(data, aes(x = subcontinent, y = net_migr_rate, fill = continent)) +
  geom_boxplot() +
  facet_grid(cols = vars(continent), scales = "free_x", space = "free_x") +
  labs(title = "Boxplots of net migration rates") +
  ylab("net migration rate") +
  theme(legend.position = "none",
        axis.text.x = element_text(vjust = 1, hjust = 1, angle = 40))
```

**Conclusion:**

Under the assumption that migration mostly happens among geographical
neighbors, we can now for example see a migration pattern from Latin
America and the Caribbean to Northern America, although the continents
total migration is roughly centered. Also Australia and New Zealand is
the only subcontinent with a positive net migration in Oceania.

## f. Median net migration rate per subcontinent.

The plot in task e. shows the distribution of the net migration rate for
each subcontinent. Here you will work on visualizing only one summary
statistic, namely the median.

For each subcontinent, calculate the median net migration rate. Then
create a plot which contains the sub-regions on the y-axis and the
median net migration rate on the x-axis.

- As geoms use points.

- Color the points by continent -- use a colorblind friendly palette
  (see e.g.,
  [here](http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/)).

- Rename the axes.

- Using `fct_reorder` from the **forcats** package, arrange the levels
  of subcontinent such that in the plot the lowest (bottom) subcontinent
  contains the lowest median net migration rate and the upper most
  region contains the highest median net migration rate.

- Comment on the plot. E.g., what are the regions with the most influx?
  What are the regions with the most outflux?

```{r}
data %>% group_by(subcontinent, continent) %>% 
  summarise(median_net_migr_rate = median(net_migr_rate, na.rm = TRUE),.groups = "drop") %>%
  mutate(subcontinent = fct_reorder(subcontinent, median_net_migr_rate)) %>%
ggplot(aes(x = median_net_migr_rate, y = subcontinent, color = continent)) +
  geom_point(size = 2.5) +
  labs(title = "Median net migration rate per subcontinent") +
  xlab("median net migration rate") +
  scale_color_viridis_d() # use viridis palette for color-blind suitability

```
**Conclusion:**

Australia and New Zealand, Western Europe, Northern Europe and Northern America have the highest median influx, while Micronesia and Polynesia the highest median outflux.

## g. Median youth unemployment rate per subcontinent

For each subcontinent, calculate the median youth unemployment rate.
Then create a plot which contains the sub-regions on the y-axis and the
median unemployment rate on the x-axis.

- Use a black and white theme (`?theme_bw()`)

- As geoms use bars. (hint: pay attention to the statistical
  transformation taking place in `geom_bar()` -- look into argument
  `stat="identity"`)

- Color the bars by continent -- use a colorblind friendly palette.

- Make the bars transparent (use `alpha = 0.7`).

- Rename the axes.

- Using `fct_reorder` from the **forcats** package, arrange the levels
  of subcontinent such that in the plot the lowest (bottom) subcontinent
  contains the lowest median youth unemployment rate and the upper most
  region contains the highest median youth unemployment rate.

- Comment on the plot. E.g., what are the regions with the highest vs
  lowest youth unemployment rate?

```{r}
data %>% group_by(subcontinent, continent) %>% 
  summarise(median_youth_unempl_rate = median(youth_unempl_rate, na.rm = TRUE),.groups = "drop") %>%
  mutate(subcontinent = fct_reorder(subcontinent, median_youth_unempl_rate)) %>%
ggplot(aes(x = median_youth_unempl_rate, y = subcontinent, fill = continent)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  labs(title = "Median youth unemployment rate per subcontinent") +
  xlab("median youth unemployment rate") +
  theme_bw() +
  scale_color_viridis_d() # use viridis palette for color-blind suitability

```
**Conclusion:**

The *stat = "identity"* argument in *geom_bar()* makes it behave like geom_col(), where instead of counting the number of categories in the categorical variable, it takes a second numeric variable, which it maps onto the bars length.

While Central, Eastern and South-eastern Asia have very low median youth unemployment rates, Northern Africa and Polynesia have the largest median youth unemployment rates.

## h. Median youth unemployment rate per subcontinent -- with error bars

The value displayed in the barplot in g. is the result of an
aggregation, so it might be useful to also plot error bars, to have a
general idea on how precise the median unemployment is. This can be
achieved by plotting the error bars which reflect the standard deviation
or the interquartile range of the variable in each of the subcontinents.

Repeat the plot from Task g. but include also error bars which reflect
the 25% and 75% quantiles. You can use `geom_errorbar` in **ggplot2**.

```{r}
data %>% group_by(subcontinent, continent) %>% 
  summarise(median_youth_unempl_rate = median(youth_unempl_rate, na.rm = TRUE),
            Q1 = quantile(youth_unempl_rate, 0.25, na.rm = TRUE),
            Q3 = quantile(youth_unempl_rate, 0.75, na.rm = TRUE), .groups = "drop") %>%
  mutate(subcontinent = fct_reorder(subcontinent, median_youth_unempl_rate)) %>%
ggplot(aes(x = median_youth_unempl_rate, y = subcontinent, fill = continent)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_errorbar(aes(xmin= Q1, xmax = Q3), width = 0.5) +
  labs(title = "Median youth unemployment rate per subcontinent") +
  xlab("median youth unemployment rate") +
  theme_bw() +
  scale_color_viridis_d() # use viridis palette for color-blind suitability

```

## i. Relationship between education expenditure and net migration rate

Using **ggplot2**, create a plot showing the relationship between
education expenditure and net migration rate.

- Color the geoms based on the income status.

- Add a regression line for each development status (using
  `geom_smooth()`).

Comment on the plot. Do you see any relationship between the two
variables? Do you see any difference among the income levels?

```{r}
ggplot(data, aes(x = expenditure, y = net_migr_rate, color = income_status)) +
  geom_jitter(alpha = 0.6) +
  ylim(-20,20) +
  geom_smooth(method = "lm", se = FALSE, fullrange = TRUE, alpha =0.6, linewidth = 0.4) +
  labs(title = "Relationship between education expenditure and net migration rate",
       x = "Education expenditure (% of GDP)",
       y = "Net migration rate") +
  theme_bw() 
```
We set the y limits to -20 and 20 to make the plot more readable (Lebanon and Syria are extreme outliers as previously seen in the boxplots). There seems to be a slight downtrend in net migration as education expenditure increases, most noticably for high-income countries. However, the relationship is not very strong and there is a lot of variability in net migration rate across all levels of education expenditure. The regression lines for the different income levels are relatively close to each other, suggesting that the relationship between education expenditure and net migration rate does not differ substantially among the income levels.


## j. Relationship between youth unemployment and net migration rate

Create a plot as in Task i. but for youth unemployment and net migration
rate. Comment briefly.

## k. Merging population data

Go online and find a data set which contains the 2020 population for the
countries of the world together with ISO codes.

- Download this data and merge it to the dataset you are working on in
  this case study using a left join. (A possible source: [World
  Bank](https://data.worldbank.org/indicator/SP.POP.TOTL?end=2020&start=2020)))

- Inspect the data and check whether the join worked well.

First we import the dataset and convert it to a tidy format:
```{r}
population_data <- read_excel("population_data.xls")

population_data_20 <- population_data %>%
                       pivot_longer(5:70, names_to = "year", values_to = "population") %>%
                        filter(year == 2020) %>%
                         select(ISO3    = `Country Code`,
                                population2020 = population)

glimpse(population_data_20)
```
Next we perform the left join with our current data:
```{r}
data <- left_join(data, population_data_20, by = "ISO3")

glimpse(data)
```
We have successfully added the population2020 column for all columns that were already present in our data.

## l. Scatterplot of education expenditure and net migration rate in Europe

Make a scatterplot of education expenditure and net migration rate for
the countries of Europe.

- Scale the size of the points according to each country's population.

- For better visibility, use a transparency of `alpha=0.7`.

- Remove the legend.

- Comment on the plot.

```{r}
data %>% filter(continent == "Europe" & !is.na(expenditure) ) %>%
          ggplot(aes(x = net_migr_rate, y = expenditure, size = population2020)) +
          geom_point(alpha = 0.7, color = "darkblue") +
          labs(title = "Net migration vs. education expenditure in Europe",
               x = "Net migration rate",
               y = "Education expenditure") +
          guides(size = "none") # remove the legend for size
```
**Conclusion:**

One visible trend is that net migration rate increases with education expenditure. All the countries with the most education spending lie in the positive midfield of net migration rate. The most populated countries in Europe all seem to have a population expenditure between 4 and 6. 

## m. Interactive plot

On the merged data set from Task k., using function `ggplotly` from
package **plotly**\
re-create the scatterplot in Task l., but this time for all countries.
Color the points according to their continent.

When hovering over the points the name of the country, the values for
education expenditure, net migration rate, and population should be
shown. (Hint: use the aesthetic `text = Country`. In `ggplotly` use the
argument `tooltip = c("text", "x", "y", "size")`).

**Hint:** If you are creating a .pdf file, keep in mind that PDF output
is static and does not support HTML widgets. To avoid build errors, you
can wrap interactive plots in a conditional check or set the chunk
option `eval=FALSE`. Alternatively, you can use the **webshot** package
to automatically convert interactive elements into static images during
the knit process.

## n. Parallel coordinate plot

In **parallel coordinate plots** each observation or data point is
depicted as a line traversing a series of parallel axes, corresponding
to a specific variable or dimension. It is often used for identifying
clusters in the data.

One can create such a plot using the **GGally** R package. You should
create such a plot where you look at the three main variables in the
data set: education expenditure, youth unemployment rate and net
migration rate. Color the lines based on the income status. Briefly
comment.

## o. World map visualisation

Create a world map of the education expenditure per country. You can use
the vignette
<https://cran.r-project.org/web/packages/rworldmap/vignettes/rworldmap.pdf>
to find how to do this in R. Alternatively, you can use other packages
(such as **ggplot2**, **sf** and **rnaturalearthdata**) to create a map.
